---
title: "Portfolio metrics and watershed characteristics"
author: "M. Snyder"
date: "Mar. 17, 2021"
output: html_document
editor_options: 
  chunk_output_type: console
chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
library(vegan)
library(BiodiversityR)
library(knitr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(MASS)
library(data.table)
library(factoextra)

```

### Analysis excludes lake populations
```{r import, include=FALSE, echo=FALSE}
setwd("C:/Users/snyde/git_repo/occ_portfolio/Portfolio_metrics/Results")

# import portfolio measurements
metrics <- read.csv("C:/Users/snyde/Dropbox/Resiliency/Marcia/Data/Portfolio_metrics/DD_metric_nolakes031721.csv")
# drop first column
metrics <- metrics[,-1]

# import watershed metrics
CohoDataSummary <- read_csv("C:/Users/snyde/Dropbox/Resiliency/Marcia/Data/watershed_metrics021621.csv")
# drop first column
CohoDataSummary <- CohoDataSummary[,-1]
# drop lake data
CohoDataSummary <- filter(CohoDataSummary, PopulationLong != "Siltcoos River (Lake)" & PopulationLong != "Tenmile Lake" & PopulationLong != "Tahkenitch Lake")

```


```{r correlation, echo=FALSE}
landscape_raw<-CohoDataSummary[,4:dim(CohoDataSummary)[2]] # drop first 3 columns because character
cor_matrix<-cor(landscape_raw, method="spearman")
#write.csv(cor_matrix,"cor_matrix.csv")
```


Since, TotalLength_m is representing watershed size, the other variables are in terms of percent of the watershed area or length rather than area or length. Variables dropped are listed below.

```{r drop vars, echo=TRUE}
# subset metrics

landscape<-dplyr::select(landscape_raw, -HighCritHab_m, -HighCritHab_p,-MeanElev_m,
     -HighIP_m, -Watershed_sqkm, -HighIP_p, -Public_sqkm,
     -lake_area_km2, -est_area_km2, -wetland_area_km2,
      -LowGradient_m, -LogDriveLength_m, -Latitude)
```

Before doing the PCA, I scale the variables by subtracting the mean and dividing by the standard deviation rather than adding one and taking the log as Griffiths et al. 2014 did because negative numbers can't be logged.

```{r transform watershed covars, include=FALSE, echo=FALSE}

# need to scale and center data
landscape_z<-scale(landscape, scale=TRUE, center=TRUE)
```


## PCA

```{r prcomp pca, echo = FALSE}
# pca
pca_test<-prcomp(landscape_z, scale=FALSE)
# loadings can be accessed with $rotation
#pca_test$rotation # pca loading value used in linear equation to calculate score
#pca_test$x # pca scores
# test signifciance of eigen values a different way
#ordi.monte(data4,ord='pca',dim=3)
#ordi.scree(pca_test,ord='pca')
screeplot(pca_test)
fviz_eig(pca_test)

#fviz_pca_ind(pca_test,
#             col.ind = "cos2", # Color by the quality of representation
#             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#             repel = TRUE     # Avoid text overlapping
#             )

fviz_pca_biplot(pca_test, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )

eig.val <- get_eigenvalue(pca_test)
eig.val

get_pca_var(pca_test)
res.var <- get_pca_var(pca_test)
#res.var$contrib        # Contributions to the PCs
#res.var$coord          # Coordinates
#res.var$cos2     

```

## PCA interpretation
Based on eigen values values I use principal components 1-3 in the rest of the script. I think there is some support for PC3 which explains ~6-7% of the variance each but I think a 10% of variance explained is a more robust cut-off.

```{r pca variable loadings}
# eigen vectors
# weights used in the principal component = pca_test$rotation 
# pca.eigenvec(pca_test,dim=3,digits=3,cutoff=.20) #
# # structure coefficients
# pca.structure(pca_test,landscape_z,dim=3,cutoff=.10) # 
# # variance accounted for in total by retained principal components by adding up all squared structure coefficients of retained principal components
# pca.communality(pca_test,landscape_z,dim=3)
```

```{r}
# library(ggbiplot)
# ggbiplot::ggbiplot(pca_test)
```


## Biplot of PCA.
```{r biplot, echo=FALSE, eval=TRUE, include=TRUE}
#biplot(pca_test)
landscape_z_df<-as.data.frame(landscape_z) # convert matrix to dataframe
data5<-cbind(CohoDataSummary$Population, landscape_z_df) # add population column
#colnames(data5)[1]<-"Population" # rename population column
pca_4_plot<-prcomp(data5[,2:dim(data5)[2]], scale=FALSE) 
biplot(pca_4_plot, xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
biplot(pca_4_plot,  xlabs=rep("", 18),  xlim=c(-0.5,0.5), cex=0.75)
# subset biplot
PC1<-pca_test$rotation[,1]
PC2<-pca_test$rotation[,2]
PC3<-pca_test$rotation[,3]

loadings<-as.data.frame(cbind(PC1, PC2, PC3))
write.csv(loadings, "C:/Users/snyde/Dropbox/Resiliency/Marcia/Data/Portfolio_metrics/loadings_nolakes.csv")
```

## biplot loadings >0.2 PC2
```{r loadings subset1, echo=FALSE, fig.height=5, fig.width=5}
# biplot loadings >0.2 PC2
loadings_20p<-loadings[which(abs(loadings$PC2) >0.2), ]
loadings_20p_pc3<-loadings[which(abs(loadings$PC3) >0.2), ]
# biplot loadings >0.2 PC2 or PC3
loadings_20p2<-loadings[which(abs(loadings$PC2) >0.2 | abs(loadings$PC3) >0.2), ]
# loading >0.2 PC1
loadings_20p_PC1<-loadings[which(abs(loadings$PC1) >0.2), ]

```


```{r more biplots }
plot(PC2~PC1, data=loadings_20p, type='n', xlim=c(-0.4, 0.4), ylim=c(-0.4, 0.4))
text(loadings_20p$PC1,loadings_20p$PC2,label=row.names(loadings_20p),col='blue', pos=c(1,1,1,1,1,1,1,2,4,1 ))
```

## biplot loadings >0.2 on PC2 or PC3
```{r biplot subset2, echo=FALSE, fig.height=5, fig.width=5}


png("score_biplot_GT2_PC2PC3.png", units="in", height=5, width=5, res=96)
pc2_pc2_gt0.2 <- plot(PC3~PC2, data=loadings_20p2, type='n', xlim=c(-0.5, 0.4), ylim=c(-0.4, 0.4))
text(loadings_20p2$PC2,loadings_20p2$PC3,label=row.names(loadings_20p2),col='blue', pos=c(1,1,1,4,1,1,1,4,1,2,1,4,1,2,4,3 )) 
pc2_pc2_gt0.2
dev.off()
pc2_pc2_gt0.2
### 1 = below
### 2 = left
### 3 = above
### 4 = right
```

## Biplot of with subset of variables that have > 0.2 loadings weight on PC1 or PC2. 
```{r biplot subset3, echo=FALSE, fig.height=8, fig.width=8}
pca_4_plot_subset<-pca_4_plot
pca_4_plot_subset$rotation<-pca_4_plot_subset$rotation[which(abs(pca_4_plot_subset$rotation[,2])>0.2 | abs(pca_4_plot_subset$rotation[,1])>0.2)  ,]
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
png("score_biplot_subset_nolakes.png", units="in", height=8, width=8, res=96)
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
dev.off()

pdf("score_biplot_subset.pdf",  height=8, width=8)
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
dev.off()
```

## Biplot of with subset of variables that have > 0.2 loadings weight on PC2 or PC3. 
```{r biplot subset4, echo=FALSE, fig.height=8, fig.width=8}
pca_4_plot_subset<-pca_4_plot
pca_4_plot_subset$rotation<-pca_4_plot_subset$rotation[which(abs(pca_4_plot_subset$rotation[,3])>0.2 | abs(pca_4_plot_subset$rotation[,2])>0.2)  ,]
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
png("score_biplot_subset_nolakes.png", units="in", height=8, width=8, res=96)
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
dev.off()

pdf("score_biplot_subset_pc3_pc2_nolakes.pdf",  height=8, width=8)
biplot(pca_4_plot_subset,  xlabs=data5[,1],  xlim=c(-0.5,0.5), cex=0.75)
dev.off()
```


```{r barplot pc1, echo=FALSE}
# reverse direction of sign so that positive value indicates positive correlation with Sharpes
loadings_20p_PC1$PC1_neg<-loadings_20p_PC1$PC1*1
# barplot with ggplot of loadings on PC2
p <-ggplot(loadings_20p_PC1, aes(x=rownames(loadings_20p_PC1), y=PC1_neg))
p +geom_bar(stat = "identity", fill="steelblue") +
  theme_minimal() + 
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14), 
        axis.title.y = element_text(face="bold", colour="#990000", size=16)) + scale_y_continuous(name="PC1 >abs(0.2)")
```

```{r barplot pc3 and pc2, echo=FALSE}
# reverse direction of sign so that positive value indicates positive correlation with Sharpes
loadings_20p2$PC2_neg<-loadings_20p2$PC2*1
# barplot with ggplot of loadings on PC2
p <-ggplot(loadings_20p2, aes(x=rownames(loadings_20p2), y=PC2_neg))
p +geom_bar(stat = "identity", fill="steelblue") +
  theme_minimal() + 
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14), 
        axis.title.y = element_text(face="bold", colour="#990000", size=16)) + scale_y_continuous(name="PC2 or PC3 Loadings >abs(0.2)")
```

## Barplot including variables with loadings on PC2 > 0.2
```{r barplot pc2, echo=FALSE}
loadings_20p$PC2_neg<-loadings_20p$PC2 * 1
# barplot with ggplot of loadings on PC2
p <-ggplot(loadings_20p, aes(x=rownames(loadings_20p), y=PC2_neg))
p +geom_bar(stat = "identity", fill="steelblue") +
  theme_minimal() + 
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14), 
        axis.title.y = element_text(face="bold", colour="#990000", size=16)) +  scale_y_continuous(name="PC2 Loadings >abs(0.2)")
```


## Barplot including variables with loadings on PC3 > 0.2
```{r barplot PC3, echo=FALSE}
loadings_20p_pc3$PC3_neg<-loadings_20p_pc3$PC3 * 1
# barplot with ggplot of loadings on PC2
p <-ggplot(loadings_20p_pc3, aes(x=rownames(loadings_20p_pc3), y=PC3_neg))
p +geom_bar(stat = "identity", fill="steelblue") +
  theme_minimal() + 
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14), 
        axis.title.y = element_text(face="bold", colour="#990000", size=16))+  scale_y_continuous(name="PC3 Loadings >abs(0.2)")
```

Convert principal components to scores for each watershed and join tables with CV, DD,with watershed characteristics, and PCA scores.
```{r pca scores join2, echo=FALSE}
# grab principal component scores for significant components
pc1<-pca_test$x[1:18,1]
pc2<-pca_test$x[1:18,2]
pc3<-pca_test$x[1:18,3]
#class(pc1)
data_pca<-cbind(data5,pc1, pc2, pc3)
#View(OC_COHO_SharpeRatio_CV)
data_pca2<-merge(metrics, data_pca, by.x="pops_names", by.y="CohoDataSummary$Population")
write.csv(data_pca2, "PCA_metrics_scores_031721_nolakes.csv")

```

```{r pca loadings join, echo=FALSE}
# grab principal component scores for significant components
pc1_loadings<-pca_test$rotation[1:28,1]
pc2_loadings<-pca_test$rotation[1:28,2]
pc3_loadings<-pca_test$rotation[1:28,3]
#class(pc1)
vars1<-names(data5)
data_pca_loadings<-cbind(vars1[-1],pc1_loadings, pc2_loadings, pc3_loadings)
#View(OC_COHO_SharpeRatio_CV)
write.csv(data_pca_loadings, "PCA_metrics_loadings_031721_nolakes.csv")

```


### Test for correlations again between significant PCA components with CV 
```{r pca correlation w CV and Sharpes2, echo=FALSE}
#names(data_pca2)

CV_pca1<-cor(data_pca2$CVCurrent_LOO_V , data_pca2$pc1,method= "spearman")
CV_pca2<-cor(data_pca2$CVCurrent_LOO_V , data_pca2$pc2, method= "spearman")
CV_pca3<-cor(data_pca2$CVCurrent_LOO_V , data_pca2$pc3, method= "spearman")

```


```{r table correlation coefficients2, echo=FALSE}

digits=2
labels<-c( "PC1", "PC2","PC3")
results1<-c(CV_pca1, CV_pca2,CV_pca3)

results3<-cbind(labels, results1)
out_table<-kable(results3, digits=getOption("digits"), col.names=c("Principal component","CVcurrent"))
out_table
```

### Test for correlations again between significant PCA components with DD 
```{r correlation betweeen DD and PCA scores, echo=FALSE}
#names(data_pca2)

DD_pca1<-cor(data_pca2$DD , data_pca2$pc1,method= "spearman")
DD_pca2<-cor(data_pca2$DD , data_pca2$pc2, method= "spearman")
DD_pca3<-cor(data_pca2$DD , data_pca2$pc3, method= "spearman")

```


```{r table correlation coefficients DD, echo=FALSE}

digits=2
labels<-c( "PC1", "PC2","PC3")
results1<-c(DD_pca1, DD_pca2,DD_pca3)

results3<-cbind(labels, results1)
out_table<-kable(results3, digits=getOption("digits"), col.names=c("Principal component","DD"))
out_table
```

```{r barplot diversity deficit}
ggplot(metrics, aes(x=pops_names, y = DD))+
  geom_point()+
  theme_classic()+
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14))+
  xlab(" ")

ggplot(metrics, aes(x=pops_names, y = CVCurrent_LOO_V))+
  geom_point()+
  theme_classic()+
  theme(axis.text.x  = element_text(angle=45, vjust=0.5, size=14))+
  xlab(" ")+
  ylab("CV")
  
```



```{r PC plots, echo=FALSE, include=TRUE, eval=TRUE, fig.height=5, fig.width=9}
# par(mfrow=c(3,2), mai = c(0.5, 0.5, 0.2, 0.2))
# plot(CVcurrent_LOO_V~pc1, data=data_pca2, xlab="PC1")
# 
# plot(CVcurrent_LOO_V~pc2, data=data_pca2, xlab="PC2")
# 
# plot(CVcurrent_LOO_V~pc3, data=data_pca2, xlab="PC3")
```


```{r plots, echo=FALSE, fig.height=10, fig.width=7}
#par(mfrow=c(5,2)) #, mai = c(1.0, 1.0, 0.2, 0.2) # bottom, left, top, right
# par(mfrow=c(5,2), mai = c(0.5, 1.0, 0.2, 0.2))
# plot(CVcurrent_LOO_V~MeanAnnualPrecip_mm, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~MeanBFI, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~MeanAugtemps_c, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~Sedimentary_p, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~Latitude, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~CVAugtemps, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~pHatchery_Mean, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~lake_p, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~lake_highCrit_ratio, data=data_pca2)
# 
# plot(CVcurrent_LOO_V~LowGradient_p, data=data_pca2)

```

